---
title: "Analysis of US Weather and The Effects of Major Events on Population Health and Economic Status"
output: 
  html_document:
    keep_md: true
---

###Synopsis
This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. It mainly looks into the types of events that are most harmful with respect to population health (i.e. fatalities and injuries). In addition, it highlights the events affecting properties and crops resulting in severe economic consequences. The reported results indicate the top-ten events with negative consequences in each case and provide plots to show the calculated values. The most harmful events are also listed separately and cumulatively in tables, whether in the case of health or economic consequences.

###Data Processing 

```{r Library_Call, results='hide', message=FALSE, warning=FALSE}
#calling libraries
setwd("G:/MOOC_Courses/Data Science Toolbox/5-Reproducible Research/Week4/Assign2/Weather_Analysis_Consequences")
library(dplyr)
library(ggplot2)
library(tidyr)

```

####-Downloading and reading data

```{r Data_Download, cache=TRUE}
#download and unzip data
data_url<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

if(!file.exists("weather.csv.bz2")) 
{download.file(data_url, destfile = "weather.csv.bz2")}

#read data
storm_data<-read.csv("weather.csv.bz2",header = T, stringsAsFactors = F)

```

####-Reading and editing the data

```{r lookup table, cache=TRUE}
#create a lookup table

lut<-c("0"=0,"1"=1,"2"=2,"3"=3,"4"=4,"5"=5,"6"=6,"7"=7,"8"=8,"h"=2,"H"=2,"K"=3,"m"=6,"M"=6,"B"=9)
# 
# #replace the values in PROPDMGEXP & CROPDMGEXP columns with the corresponding values in the lookup table
storm_data$PROPDMGEXP<-(lut[storm_data$PROPDMGEXP])
storm_data$CROPDMGEXP<-(lut[storm_data$CROPDMGEXP])

```
###Results

####-Types of events that are most harmful with respect to population health

The following results and plots show that tornados are the worst effects in terms of negative consequences on health as they cause thousands of injuries and fatalities. Excessive heat, TSTM Wind and floods come next if we look on their cumulative effect. However their severity differs with respect to injuries and fatalities. And it is clear that the number of injuries dominates when both injuries and fatalities are summed up.

```{r}
#calculate the number of fatalities and injuries corresponding to each event type
E1<-storm_data %>%
        select(EVTYPE,FATALITIES, INJURIES) %>%
        group_by(EVTYPE) %>%
        summarise(Fatalities=sum(FATALITIES),Injuries=sum(INJURIES)) %>%
        mutate(Total=(Fatalities+Injuries))

#order the events based on the cumulative effect (fatalities+injuries)
E1t<-E1[order(E1$Total,decreasing = T),]

#Extract and order the top events corresponding to fatalities
E1f<-select(E1,EVTYPE,Fatalities)[order(E1$Fatalities,decreasing = T),]

#Extract and order the top corresponding to injuries
E1i<-select(E1,EVTYPE,Injuries)[order(E1$Injuries,decreasing = T),]
```

-Top tep ten events causing the most fatalities+injuries cumulatively:
```{r echo=FALSE}
head(E1t,10)
```
-Top tep ten events causing the most fatalities:
```{r echo=FALSE}
head(E1f,10)
```
-Top tep ten events causing the most injuries:
```{r echo=FALSE}
head(E1i,10)
```

####Plotting data events that are most harmful with respect to population health
```{r plotting health effects }
#fatalities Vs. events
pf<-ggplot(data=head(E1f,10),aes(x=reorder(EVTYPE,-Fatalities) ,y=Fatalities))+
        geom_bar(stat="identity")+
        scale_x_discrete(name="Event Type")+
        scale_y_continuous(name="Fatalities")+
        labs(title="Top 10 Events Causing Fatalaties")+
        theme(axis.text.x=element_text(angle = 90,vjust=1))

#Injuries Vs. events
pi<-ggplot(data=head(E1i,10),aes(x=reorder(EVTYPE,-Injuries) ,y=Injuries))+
        geom_bar(stat="identity")+
        scale_x_discrete(name="Event Type")+
        scale_y_continuous(name="Injuries")+
        labs(title="Top 10 events causing injuries")+
        theme(axis.text.x=element_text(angle = 90,vjust=1))

#gather data to use fatalities/injuries for coloring the bars
kk<-gather(head(E1t,10), value="Total_by_Type",key="type",Fatalities,Injuries,na.rm = T)

#Fatalities+Injuries Vs. events
pp1<-ggplot(data=kk,aes(x=reorder(EVTYPE,-Total_by_Type),y=Total_by_Type,fill=type))+
        geom_bar(stat="identity")+
        scale_x_discrete(name="Event Type")+
        scale_y_continuous(name="Fatalities+Injuries")+
        labs(title="Top 10 events with negative consequences on health (Injuries+Fatalities)")+
        theme(axis.text.x=element_text(angle = 90,vjust=1))+
        scale_fill_grey()
       
```

```{r printing plot_1_total,echo=F}
pp1

```

```{r printing plot_1_fatalities,echo=F}
pf

```

```{r printing plot_2_injuries,echo=F}
pi
```


####-Types of events that have the greatest economic consequences

The following results and plots show that floods cause the highest losses in terms of properties, then comes the hurricanes/typhoons, tornadoes and storm surges. On the other hand droughts cause the highest crop damages, then comes the floods, river floods and ice storms. The results also indicate that the economic losses due to property damages are higher than the losses due to crop damages. Consuquently property damages' losses dominate the sum of both losses for most of the top ten cumulative losses.


```{r}
#calculate the number property and crop damage corresponding to each event type
E2<-storm_data %>%
        select(EVTYPE,PROPDMG,PROPDMGEXP, CROPDMG,CROPDMGEXP) %>%
        group_by(EVTYPE) %>%
        summarise(Property_damage=sum(PROPDMG*10^PROPDMGEXP,na.rm=T),
                Crop_damage=sum(CROPDMG*10^CROPDMGEXP,na.rm=T)) %>%
        mutate(Total_Dmg=Property_damage+Crop_damage)

#order the events based on the cumulative effect (property damage+crop damage)
E2t<-E2[order(E2$Total_Dmg,decreasing=T),]

#Extract and order the top events corresponding to property damage
E2p<-select(E2,EVTYPE,Property_damage)[order(E2$Property_damage,decreasing = T),]

#Extract and order the top events corresponding to crop damage
E2c<-select(E2,EVTYPE,Crop_damage)[order(E2$Crop_damage,decreasing = T),]
```


-Top tep ten events causing the highest economic losses cumulatively (property damage+crop damage):
```{r echo=F}
head(E2t,10)
```
-Top tep ten events causing the higest property damages:
```{r echo=F}
head(E2p,10)
```
-Top tep ten events causing the higest crop damages:
```{r echo=F}
head(E2c,10)
```

####Plotting events that have the greatest economic consequences

```{r}
#Property Damage Vs. events
ppd<-ggplot(data=head(E2p,10),aes(x=reorder(EVTYPE,-Property_damage) ,y=Property_damage/10^9))+
        geom_bar(stat="identity")+
        scale_x_discrete(name="Event Type")+
        scale_y_continuous(name="Property Damage (*10^9)")+
        labs(title="Top 10 Events Causing highest property damage")+
        theme(axis.text.x=element_text(angle = 90,vjust=1))

#Crop Damage Vs.events
pcd<-ggplot(data=head(E2c,10),aes(x=reorder(EVTYPE,-Crop_damage) ,y=Crop_damage/10^9))+
        geom_bar(stat="identity")+
        scale_x_discrete(name="Event Type")+
        scale_y_continuous(name="Crop Damage (*10^9)")+
        labs(title="Top 10 Events Causing highest crop damage")+
        theme(axis.text.x=element_text(angle = 90,vjust=1))

#gather data to use Property_damage,Crop_damage for coloring the bars
mm<-gather(head(E2t,10), value="Total_by_Type",key="type",Property_damage,Crop_damage,na.rm = T)

#Property Damage+Crop Damage Vs. events
pp2<-ggplot(data=mm,aes(x=reorder(EVTYPE,-Total_by_Type),y=Total_by_Type/10^9,fill=type))+
        geom_bar(stat="identity")+
        scale_x_discrete(name="Event Type")+
        scale_y_continuous(name="Property Damage+Crop Damage (*10^9)")+
        labs(title="Top 10 events with negative consequences on health (Injuries+Fatalities)")+
        theme(axis.text.x=element_text(angle = 90,vjust=1))+
        scale_fill_grey()
```

```{r printing plot_2_total,echo=F}
pp2

```

```{r printing plot_1_properties,echo=F}
ppd

```

```{r printing plot_2_crops,echo=F}
pcd
```

In conclusion, the weather events thet cause th highest public health and economic problems for communities and municipalities could be highlighted and the losses were reported per event. This could be used to allocate resources for proactive actions in order to reduce the losses whenever possible. 
